Optimal tensor denoising and completion based on ordinal observations

### 2020-01-24 ICML/miaoyan/ordinalT.tex
update

\begin{table*}[http]
\centerline{\includegraphics[width=16cm]{compare.pdf}}
\vspace{-.4cm}
\caption{Comparison of various low-rank estimation methods based on categorical observations. For an order-$K$ dimensional-$(d,\ldots,d)$ ordinal tensor with $L$-level observations, we report the error bound in the recovered signal (for tensor denoising) and required sample complexity (for tensor completion) as functions of tensor dimension. {\color{red}(font in the table seems odd...)}}~\label{tab:compare}
\end{table*}

@@

\begin{assumption}\label{ass:link}
The link function is assumed to satisfy:
\vspace{-.4cm}
\begin{enumerate}
\item $f(\theta)$ is strictly increasing, strictly log-concave, and twice-differentiable in $\theta\in\mathbb{R}/\{0\}$.
\vspace{-.2cm}
\item $f'(\theta)$ is unimodal and symmetric with respect to $\theta=0$. 
\end{enumerate}
\end{assumption}

@@

We define a few key quantities that will be used in our theory. Let $g_\ell=f(\theta+b_\ell)-f(\theta+b_{\ell-1})$ for all $\ell\in[L]$, and
\begin{align}\label{eq:regular}
A_\alpha &= \min_{\ell\in[L], |\theta|\leq \alpha} g_\ell(\theta),\quad  U_\alpha=\max_{\ell\in[L], |\theta|\leq \alpha } {\dot{g}_\ell(\theta)\over g_\ell(\theta)},\\
 L_\alpha &= \min_{\ell\in[L], |\theta|\leq \alpha }\left[{\dot{g}_\ell^2 (\theta)\over g_\ell^2(\theta)} -{\ddot{g}_\ell(\theta)\over g_\ell(\theta)}\right],
\end{align}
where $\dot{g}(\theta)=dg(\theta)/d\theta$, and $\alpha$ is the entrywise bound of $\Theta$. In view of equation~\eqref{eq:latent}, these quantities characterize the geometry (such as flatness and convexity) of the latent noise distribution. Under the assumption~\ref{ass:link}, all these quantities are strictly positive and independent of the tensor dimension.

@@

A similar conclusion can be obtained for the prediction error, measured in KL divergence, between the two generative distributions. 
\begin{cor}[Prediction error]~\label{cor:prediction}
Assume the same set-up as in Theorem~\ref{thm:rate}. Let $\mathbb{P}_{\tY}$ and $\hat{\mathbb{P}}_{\tY}$ denote the distributions generating the $L$-level ($L\geq 2$) ordinal tensor $\tY$, given the true parameter $\Theta$ and its estimator $\hat \Theta$, respectively. Then, with very high probability, 
\begin{equation}\label{eq:KLrate}
\text{KL}(\mathbb{P}_{\tY} || \hat{\mathbb{P}}_{\tY}) \leq  {c_2 U^2_\alpha r_{\max}^{K-1}  \over L^2_\alpha } {(4L-6)\dot{f}^2(0)  \over A_\alpha} {\sum_kd_k\over  \prod_k d_k},
\end{equation}
where $c_1, c_2>0$ are the same constants as in Theorem~\ref{thm:rate}.  
\end{cor}
To gain insight into these bounds, we consider a special setting where the dimensions are the same in all modes, i.e., $d_1=\cdots=d_K=d$. In such a case, our bound \eqref{eq:rate} reduces to
\begin{equation*}\label{eq:ours}
\text{MSE}(\hat \Theta, \trueT) \asymp d^{-(K-1)}, \quad \text{as}\ d\to \infty.
\end{equation*}
Hence, the estimator achieves consistency with polynomial convergence rate. Comparing to max norm based estimator $\tilde{\Theta}$ proposed by~\citet{ghadermarzy2018learning}, $\text{MSE}(\tilde{\Theta}, \trueT) \asymp  d^{-(K-1)/2}$, we see that our bound has a faster convergence rate. We believe the improvement comes from utilizing the exact low-rank structure of $\Theta$, whereas an scale-sensitive surrogate rank measure was employed in~\citet{ghadermarzy2018learning}.

Our bound also generalizes the previous results on ordinal matrices. The convergence rate for rank-constrained matrix denoising was $O(1/\sqrt{d})$~\citep{bhaskar2016probabilistic}, which fits into our special case when $K=2$. Furthermore, our results~\eqref{eq:rate} and~\eqref{eq:KLrate} reveal that the convergence becomes favorably as the order of tensor data increases. 
Intuitively, for tensor data analysis, we can view the number of entries, $\prod_k d_k$, as the sample size, and the number of free parameters in $\tP$ is roughly on the order of $\sum_{k}d_k$, assuming $r_{\max}=O(1)$. Compared to the matrix case, a higher tensor order implies more effective samples per parameter and thus exhibits a faster convergence rate in high dimensions.

@@

\begin{thm}[Minimax lower bound]\label{thm:minimax} 
Assume the same set-up as in Theorem~\ref{thm:rate}, and $d_{\max}=\max_k d_k \geq 8$. Let $\inf_{\hat \Theta}$ denote the infimum over all estimators $\hat \Theta\in\tP$ based on the ordinal tensor observation $\tY\in[L]^{d_1\times \cdots \times d_K}$. Then, under the model~\eqref{eq:model},
\begin{align}\label{eq:lower}
\inf_{\hat \Theta }\sup_{\trueT\in\tP} \mathbb{P}\Big\{& \text{MSE}(\hat \Theta, \trueT) \\
& \geq {1\over 256} \min\left( \alpha^2, \ { Cr_{\max}d_{\max} \over \prod_k d_k} \right) \Big\} \geq {1\over 8},
\end{align}
where $C=C(\alpha, L, f,\mb)>0$ is a constant independent of the tensor dimension and rank. 
\end{thm}

@@

Now we consider the problem of tensor completion, when only a subset of entries $\Omega$ are observed. We consider a general sampling procedure induced by $\Pi$, and assess the accuracy using the weighted squared deviation:
\begin{align}\label{eq:weighted}
\PiFnormSize{}{\Theta-\hat \Theta}^2&\stackrel{\text{def}}{=}
{1\over |\Omega|}\mathbb{E}_{\Omega\sim \Pi}\FnormSize{}{\Theta-\hat \Theta}^2\notag \\
&=\sum_{\omega\in[d_1]\times \cdots \times [d_K]} \pi_{\omega}(\Theta_{\omega}-\hat \Theta_{\omega})^2.
\end{align}

Note that the recovery error depends on the distribution. In particular, the entries with higher sampling probabilities are equipped with better accuracy guarantee, compared to the ones with lower sampling probabilities.

@@

\section{Numerical Implementation}\label{sec:algorithm}
\subsection{Alternating optimization}
\begin{figure*}[http]
\begin{center}
\includegraphics[width=16cm]{algorithm.pdf}
 \end{center}
\end{figure*}
In this section, we describe the algorithm to seek the optimizer of~\eqref{eq:objective}. The objective function $\logl(\Theta)$ is concave in $\Theta$ whenever $\tP$ is log-concave~\cite{mccullagh1980regression}. However, the feasible set $\tP$ is not a convex set, which makes the optimization~\eqref{eq:objective} a non-convex problem. One approach to handle this problem is utilizing the Tucker decomposition and converting optimization into a block-wise convex problem. Let us denote the low Tucker-rank tensor as
\begin{equation}\label{eq:Tucker}
\Theta=\tC\times_1\mM_1\times_1 \cdots \times_K \mM_K,
\end{equation}
where $\tC\in\mathbb{R}^{r_1\times \cdots \times r_K}$ is a core tensor and $\mM_k\in\mathbb{R}^{d_k\times r_k}$ are factor matrices with orthogonal columns. From~\eqref{eq:objective} and \eqref{eq:Tucker}, we have $K+2$ blocks of variables in the objective function, one for the cut-points vector $\mb$, one for the core tensor $\tC$, and $K$ for the factor matrices $\mM_k$'s. We can change the optimization problem to simple convex problem if any $N+1$ out of the $N+2$ blocks being fixed. Therefore, we can update one block at a time while other blocks being fixed, and alternate the optimization via iteration. The Algorithm 1 gives the full description.

@@

\section{Real data application}
In this section, we apply our ordinal tensor decomposition method to two real-world datasets of ordinal tensors. In the first application, we use our model to analyze an ordinal tensor consisting of structural connectivity patterns among 68 brain regions for 136 individuals from Human Connec- tome Project (HCP) (Geddes 2016). In the second application, we perform tensor completion from the data with missing values. The data tensor records the ratings from scale 1 to 5 of 42 users to 139 songs on 26 contexts (Baltrunas et al. 2011).

@@

>>>>

\begin{table*}[http]
\centerline{\includegraphics[width=16cm]{compare.pdf}}
\vspace{-.4cm}
\caption{Comparison of various low-rank estimation methods based on categorical observations. For an order-$K$ dimensional-$(d,\ldots,d)$ ordinal tensor with $L$-level observations, we report the error bound in the recovered signal (for tensor denoising) and required sample complexity (for tensor completion) as functions of tensor dimensions. {\color{red}(font in the table seems odd...)}}~\label{tab:compare}
\end{table*}

@@

\begin{assumption}\label{ass:link}
The link function is assumed to satisfy:
\vspace{-.4cm}
\begin{enumerate}
\item $f(\theta)$ is strictly increasing, strictly log-concave, and twice-differentiable in $\theta\in\mathbb{R}/\{0\}$.
\vspace{-.2cm}
\item $f'(\theta)$ is unimodal, and symmetric with respect to $\theta=0$. 
\end{enumerate}
\end{assumption}

@@

We define a few key quantities that will be used in our theory. Let $g_\ell=f(\theta+b_\ell)-f(\theta+b_{\ell-1})$ for all $\ell\in[L]$, and
\begin{align}\label{eq:regular}
A_\alpha &= \min_{\ell\in[L], |\theta|\leq \alpha} g_\ell(\theta),\quad  U_\alpha=\max_{\ell\in[L], |\theta|\leq \alpha } {\dot{g}_\ell(\theta)\over g_\ell(\theta)},\\
 L_\alpha &= \min_{\ell\in[L], |\theta|\leq \alpha }\left[{\dot{g}_\ell^2 (\theta)\over g_\ell^2(\theta)} -{\ddot{g}_\ell(\theta)\over g_\ell(\theta)}\right],
\end{align}
where $\dot{g}(\theta)=dg(\theta)/d\theta$, and $\alpha$ is the entrywise bound of $\Theta$. In view of equation~\eqref{eq:latent}, these quantities characterize the geometry (such as flatness and convexity) of the latent noise distribution. Under the Assumption~\ref{ass:link}, all these quantities are strictly positive and independent of tensor dimensions.

@@

A similar conclusion can be obtained for the prediction error, measured in Kullback-Leibler (KL) divergence, between the two categorical distributions. 
\begin{cor}[Prediction error]~\label{cor:prediction}
Assume the same set-up as in Theorem~\ref{thm:rate}. Let $\mathbb{P}_{\tY}$ and $\hat{\mathbb{P}}_{\tY}$ denote the distributions generating the $L$-level ordinal tensor $\tY$, given the true parameter $\Theta$ and its estimator $\hat \Theta$, respectively. Assume $L\geq 2$. Then, with very high probability, 
\begin{equation}\label{eq:KLrate}
\text{KL}(\mathbb{P}_{\tY} || \hat{\mathbb{P}}_{\tY}) \leq  {c_2 U^2_\alpha r_{\max}^{K-1}  \over L^2_\alpha } {(4L-6)\dot{f}^2(0)  \over A_\alpha} {\sum_kd_k\over  \prod_k d_k},
\end{equation}
where $c_1, c_2>0$ are the same constants as in Theorem~\ref{thm:rate}.  
\end{cor}
To gain insight into these bounds, we consider a special setting where the dimensions are the same in all modes, i.e., $d_1=\cdots=d_K=d$. In such a case, our bound \eqref{eq:rate} reduces to
\begin{equation*}\label{eq:ours}
\text{MSE}(\hat \Theta, \trueT) \asymp d^{-(K-1)}, \quad \text{as}\ d\to \infty.
\end{equation*}
Hence, the estimator achieves consistency with polynomial convergence rate. Comparing to max norm based estimator $\tilde{\Theta}$ proposed by~\citet{ghadermarzy2018learning}, $\text{MSE}(\tilde{\Theta}, \trueT) \asymp  d^{-(K-1)/2}$, we see that our bound has a faster convergence rate. We believe the improvement comes from utilizing the exact low-rank structure of $\Theta$, whereas an scale-sensitive surrogate rank measure was employed in~\citet{ghadermarzy2018learning}.

Our bound also generalizes the previous results on ordinal matrices. The convergence rate for rank-constrained matrix denoising was $O(1/\sqrt{d})$~\citep{bhaskar2016probabilistic}, which fits into our special case when $K=2$. Furthermore, our results~\eqref{eq:rate} and~\eqref{eq:KLrate} reveal that the convergence becomes favorably as the order of tensor data increases. 
Intuitively, for tensor data analysis, the sample size is the number of tensor entries, $\prod_k d_k$, and the number of free parameters is roughly on the order of $\sum_{k}d_k$, assuming $r_{\max}=O(1)$. A higher tensor order implies higher effective sample size per parameter, and thus exhibits a faster convergence rate in high dimensions.

@@

\begin{thm}[Minimax lower bound]\label{thm:minimax} 
Assume the same set-up as in Theorem~\ref{thm:rate}, and $d_{\max}=\max_k d_k \geq 8$. Let $\inf_{\hat \Theta}$ denote the infimum over all estimators $\hat \Theta\in\tP$ based on the ordinal tensor observation $\tY\in[L]^{d_1\times \cdots \times d_K}$. Then, under the model~\eqref{eq:model},
\begin{align}\label{eq:lower}
\inf_{\hat \Theta }\sup_{\trueT\in\tP} \mathbb{P}\Big\{& \text{MSE}(\hat \Theta, \trueT) \\
& \geq {1\over 256} \min\left( \alpha^2, \ { Cr_{\max}d_{\max} \over \prod_k d_k} \right) \Big\} \geq {1\over 8},
\end{align}
where $C=C(\alpha, L, f,\mb)>0$ is a constant independent of tensor dimensions and the rank. 
\end{thm}

@@

Now we consider the problem of tensor completion, when only a subset of entries $\Omega$ are observed. We consider a general sampling procedure induced by $\Pi$. The recovery accuracy is assessed by the weighted squared deviation:
\begin{align}\label{eq:weighted}
\PiFnormSize{}{\Theta-\hat \Theta}^2&\stackrel{\text{def}}{=}
{1\over |\Omega|}\mathbb{E}_{\Omega\sim \Pi}\FnormSize{}{\Theta-\hat \Theta}^2\notag \\
&=\sum_{\omega\in[d_1]\times \cdots \times [d_K]} \pi_{\omega}(\Theta_{\omega}-\hat \Theta_{\omega})^2.
\end{align}

Note that the recovery error depends on the distribution $\Pi$. In particular, the entries with higher sampling probabilities are equipped with better accuracy guarantee, compared to the ones with lower sampling probabilities.

@@

\section{Numerical Implementation}\label{sec:algorithm}
\subsection{Alternating optimization}

\begin{algorithm}[tb]
        \caption{Ordinal tensor decomposition }\label{alg}
        \begin{algorithmic}[]
            \STATE{\bfseries Input:}  \text{ Ordinal tensor }
            $\mathcal{Y}\in [L]^{d_1\times\cdots\times d_K}
            $,\\ \hspace{.43in} Rank $\mr\in \mathbb{N_+}^{K}$,\\
            \hspace{.43in} Entry-wise bound $\alpha\in \mathbb{R_+}$.
            \STATE{\bfseries Output:} $(\hat\Theta,\hat{\mb}) =  \argmax_{(\Theta,\mb)\in \mathcal{D}\times\mathcal{B}}  \mathcal{L}_{\mathcal{Y},\Omega}(\Theta,\mb).$
            \STATE Initialize  Core tensor $\mathcal{C}^{(0)}$,\\
            \hspace{.5in} Factor matrices $ \{\mM_1^{(0)},\cdots,\mM_K^{(0)}\}$,\\
            \hspace{.5in} Cut-off points $\mb^{(0)}$.
            \FOR{$t = 1,2,\cdots,$}
            \FOR{$k = 1,2,\cdots,K$}
            \STATE{Update $\mM_k$ while fixing other blocks.}
            \STATE $\mM_k^{(t+1)}\gets\argmax_{\mM_k\in\mathbb{R}^{d_k\times r_K}}\mathcal{L}_{\mathcal{Y},\Omega}(\mM_k)$
            \newline
             s.t. $\mnormSize{}{\Theta^{(t+1)}}\leq \alpha$, where $\Theta^{(t+1)}$ is the parameter tensor based on the current block estimates. 
            \ENDFOR
            \STATE {Update $\mathcal{C}$ while fixing other blocks: $\mathcal{C}^{(t+1)} \gets \argmax_{\mathcal{C}\in \mathbb{R}^{r_1\times\cdots\times r_K}}\mathcal{L}_{\mathcal{Y},\Omega}(\mathcal{C})$, s.t. $\mnormSize{}{\Theta^{(t+1)}}\leq \alpha$. 
               \STATE {Update $\Theta$ based on the current block estimates: $\Theta^{(t+1)} \gets \mathcal{C}^{(t+1)}\times_1\mM_1^{(t+1)}\cdots\times_K\mM_K^{(t+1)}$. 
            \STATE {Update $\mb$ while fixing $\Theta^{(t+1)}$: $\mb^{(t+1)} \gets \argmax_{\mb\in \mathbb{R}^{L-1}}\mathcal{L}_{\mathcal{Y},\Omega}\big(\Theta^{(t+1)},\mb\big)$
            \ENDFOR
            \STATE \textbf{return}
            $\Theta,\mb$
        \end{algorithmic}
    \end{algorithm}

In this section, we describe the algorithm to seek the optimizer of~\eqref{eq:objective}. The objective function $\logl(\Theta)$ is concave in $\Theta$ whenever $\tP$ is log-concave~\cite{mccullagh1980regression}. However, the feasible set $\tP$ is not a convex set, which makes the optimization~\eqref{eq:objective} a non-convex problem. One approach to handle this problem is utilizing the Tucker decomposition and converting optimization into a block-wise convex problem. Let us denote the low Tucker-rank tensor as
\begin{equation}\label{eq:Tucker}
\Theta=\tC\times_1\mM_1\times_1 \cdots \times_K \mM_K,
\end{equation}
where $\tC\in\mathbb{R}^{r_1\times \cdots \times r_K}$ is a core tensor and $\mM_k\in\mathbb{R}^{d_k\times r_k}$ are factor matrices with orthogonal columns. From~\eqref{eq:objective} and \eqref{eq:Tucker}, we have $K+2$ blocks of variables in the objective function, one for the cut-points vector $\mb$, one for the core tensor $\tC$, and $K$ for the factor matrices $\mM_k$'s. We can change the optimization problem to simple convex problem if any $N+1$ out of the $N+2$ blocks being fixed. Therefore, we can update one block at a time while other blocks being fixed, and alternate the optimization via iteration. The Algorithm 1 gives the full description.

@@

\section{Real data application}
In this section, we apply our ordinal tensor decomposition method to two real-world datasets of ordinal tensors. In the first application, we use our model to analyze an ordinal tensor consisting of structural connectivity patterns among 68 brain regions for 136 individuals from Human Connectome Project (HCP) (Geddes 2016). In the second application, we perform tensor completion from the data with missing values. The data tensor records the ratings from scale 1 to 5 of 42 users to 139 songs on 26 contexts (Baltrunas et al. 2011).

@@

\begin{table}[H]
\resizebox{\linewidth}{!}{%
\begin{tabular}{c|ccc|c}
\hline
\multirow{2}{*}{Critera}&\multicolumn{3}{c|}{Our method}&Continuous method\\
& $\hat y^{(\text{Mean})}_\omega$ & $\hat y^{(\text{Median})}_\omega$ & $\hat y^{(\text{Mode})}_\omega$ &~\cite{filipovic2015tucker}\\
\hline
MSE& &  & 0.1504&0.1604\\
MAD&  &  &0.1504&0.1600\\
MCR& &&0.1502&0.1598\\
\hline
\end{tabular}
}
\caption{Comparison of tensor denoising on HCP dataset.}
\end{table}

\subsection{InCarMusic recommendation system}
InCarMusic is a mobile application that offers music recommendation to passengers of cars based on contexts (Baltrunas et al. 2011). Our goal is to perform the tensor completion to the $42\times 139\times 26$ ordinal tensor and thereby we can offer context-specific music recommendation to users. The tensor entries consist of ordinal observations on the scale 1 to 5, the ratings of 42 users to 139 songs on 26 contexts and are encoded as NA for missing values. The number of missing values is 148,904 and the number of available values is 2,884.
We suggest three estimators $\hat y_\omega$ based on $(\hat \theta_\omega,\hat \mb)$:
\begin{itemize}
\item (Mean) $\hat y^{(\text{Mean})}_\omega=\sum_\ell \ell g_\ell(\hat \theta_\omega,\hat \mb) $;
\item (Median) $\hat y^{(\text{Median})}_\omega=\min\{\ell\in[L]\colon f_\ell(\hat \theta_\omega,\hat \mb)\geq 0.5\}$;
\item (Mode) $\hat y^{(\text{Mode})}_\omega=\argmax_\ell g_\ell(\hat \theta_\omega,\hat \mb) $
\end{itemize}
Note that, under the ordinal tensor model~\eqref{eq:model}, the estimator $\hat y^{(\text{Mean})}_\omega$ (uniformally?) minimizes $\mathbb{E}(y_\omega-\hat y_\omega)^2$, and the estimator $\hat y^{(\text{Median})}_\omega$ minimizes $\mathbb{E}|y_\omega-\hat y_\omega|$ (?). In contrast, these three estimators degenerate to a single estimator under the continuous-valued tensor decomposition model. We assess the accuracy using three metrics: Mean squared error (MSE), Mean absolute deviation (MAD), and Misclassification rate (MCR).

Results are averaged from 5 fold cross validation. {\color{red} add s.e. to the above table.}

\begin{table}[H]
\resizebox{\linewidth}{!}{%
\begin{tabular}{c|ccc|c}
\hline
\multirow{2}{*}{Critera}&\multicolumn{3}{c|}{Our method}&Continuous method\\
& $\hat y^{(\text{Mean})}_\omega$ & $\hat y^{(\text{Median})}_\omega$ & $\hat y^{(\text{Mode})}_\omega$ &~\cite{filipovic2015tucker}\\
\hline
MSE& \color{red}2.01&2.14 & 3.64&8.18\\
MAD& 1.23  &  {\color{red}1.13}&1.26&2.36\\
MCR& 0.80&0.75&{\color{red}0.54}&0.96\\
\hline
\end{tabular}
}
\caption{Comparison of tensor completion performance on InCarMusic dataset.}
\end{table}

We also attempted to run matrix ordinal mehod~\citep{bhaskar2016probabilistic} to the InCarMusic dataset. Unfortunately, the $\mb$ cannot be estimated from their method.
